---
title: "Homework 03: Use dplyr/ggplot2 to manipulate and explore data"
output: github_document
---

## Overview 

I will finish the following tasks in this assignment.

* Get the maximum and minimum of GDP per capita for all continents.

* Look at the spread of GDP per capita within the continents.

* Compute a trimmed mean of life expectancy for different years. 

* How is life expectancy changing over time on different continents?

* Report the absolute abundance of countries with low life expectancy over time by continent.

First we need to load `gapminder` dataset and `tidyverse` package and `knitr` and `reshape2` package. 

```{r}
library(gapminder)
library(tidyverse)
library(knitr)
library(reshape2)
```

## Get the maximum and minimum of GDP per capita for all continents

I group the gapminder dataset by continent and then use summarize() function where max() and min() are included to get the range of GDP per capita for all continents.

```{r}
a<-group_by(gapminder,continent)%>%
  summarize(max_gdpPercap=max(gdpPercap),min_gdpPercap=min(gdpPercap))
  kable(a)
```

I will plot maximum and minimum of GDP per capita for each continent using bar chart. 

```{r}
# merge max_gdpPercap column and min_gdpPercap into one column 
b<-melt(a,id=c("continent"))
# use bar plot to compare max_gdpPercap with min_gdpPercap
ggplot(b,aes(x=continent,y=value,fill=variable))+
  geom_bar(position="dodge",stat="identity")+scale_y_log10()+
  theme(plot.title = element_text(size=14,hjust=0.5))+
  labs(x="continent",
       y="GDP per capita - log scale",
       title="maximum and minimum of GDP per capita for all continents")
```

We can find out that there is big range bettween max GDP per capita and min GDP per capita. Oceania has smallest range perhaps since only two countries have been observed. 

## Look at the spread of GDP per capita within the continents

First, we can look at the distribution of GDP per capita for each continent from year 1952 to 2007. 

```{r}
p<-group_by(gapminder,continent)%>%
  summarize(max_gdpPercap=max(gdpPercap),
            min_gdpPercap=min(gdpPercap),
            mean_gdpPercap=mean(gdpPercap),
            median_gdpPercap=median(gdpPercap),
            sd_gdpPercap=sd(gdpPercap))
  kable(p)
```

We can use density plot to visualize the distribution of GDP per capita for each continent from year 1952 to 2007.

```{r}
ggplot(gapminder,aes(x=gdpPercap))+
  geom_density(color='red')+
  facet_wrap(~continent)+
  scale_x_continuous(breaks = seq(0, 120000, 20000),
                     labels = as.character(seq(0, 120000, 20000)),
                     limits = c(0,120000))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size=14,hjust=0.5))+
  labs(x="GDP per capita",
       y="density",
       title="Distribution of GDP per capita for each continent from 1952 to 2007")
```

Then I want to explore gdpPercap distribution for each continent in year 1952 and year 2007. 

```{r}
filter(gapminder,year %in% c(1952,2007))%>%
  group_by(continent,year)%>%
  summarize(mean_gdpPercap=mean(gdpPercap),
            median_gdpPercap=median(gdpPercap),
            sd_gdpPercap=sd(gdpPercap))%>%kable()
```

Use histogram plot to visualize the distribution of GDP per capita for each continent.

```{r}
filter(gapminder,year %in% c(1952,2007))%>%
  ggplot(aes(x=gdpPercap))+
  geom_histogram(fill='blue',alpha=0.3,bins=45)+
  facet_grid(continent~year)+
  scale_x_continuous(breaks = seq(0, 60000, 10000),
                     labels = as.character(seq(0, 60000, 10000)),
                     limits = c(0,60000))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size=14,hjust=0.5))+
  labs(x="GDP per capita",
       y="number of countries",
       title="Distribution of GDP per capita for each continent in 1952 and 2007")
```
  
## Compute a trimmed mean of life expectancy for different years

I want to compute a trimmed mean of life expectancy for each year. 

```{r}
d<-group_by(gapminder,year)%>%
  filter(lifeExp > (mean(lifeExp)-sd(lifeExp)) & 
         lifeExp < (mean(lifeExp)+sd(lifeExp)))%>% #remove data beyond 1 standard deviation
         summarize(mean_lifeExp=mean(lifeExp),
                   median_lifeExp=median(lifeExp),
                   sd_lifeExp=sd(lifeExp))
kable(d)
```

Now I want to use histogram to visualize data.

```{r}
group_by(gapminder,year)%>%
  filter(lifeExp > (mean(lifeExp)-sd(lifeExp)) & 
         lifeExp < (mean(lifeExp)+sd(lifeExp)))%>% #remove data beyond 1 standard deviation
  ggplot(aes(x=lifeExp))+
  geom_histogram(fill='red',alpha=0.5)+
  facet_wrap(~year)+
  theme(plot.title = element_text(size=14,hjust=0.5))+
  labs(x="life expectancy",
       y="number of countries",
       title="Distribution of Life Expectancy from 1952 to 2007")
```

Next I want to see the trend of trimmed mean of life expectancy over time.

```{r}
ggplot(d,aes(x=year,y=d$mean_lifeExp))+
  geom_point(color='red',alpha=0.5)+
  geom_line()+
  scale_x_continuous(limits=c(1952,2007),breaks=seq(1952,2007,5))+
  theme(plot.title = element_text(size=14,hjust=0.5))+
  labs(x="year",
       y="trimmed mean of life expectancy",
       title="Trimmed mean of life expectancy from 1952 to 2007")
```

## How is life expectancy changing over time on different continents?

I want to explore life expectancy changing over different periods for all continents.

```{r}
e<-mutate(gapminder, period = paste(substr(year,1,3),"0s",sep="",collapse = NULL))
l<-group_by(e,continent,period)%>%
  summarize(mean_lifeExp=mean(lifeExp),
            median_lifeExp=median(lifeExp),
            sd_lifeExp=sd(lifeExp))
filter(l,continent %in% c("Americas","Africa"))%>%kable()#just show data for Americas and Africa
```

Then I want to explore which country has longest life expectancy for each continent from 1952 to 2007. 

```{r}
group_by(e,continent,period)%>%
  arrange(desc(lifeExp))%>%
  summarize(country=first(country))%>%
  filter(continent=="Asia"| continent=="Africa")%>% # just show data of Asia and Africa
  kable()
```

Finally, I will plot life expectancy of all continents over years. 
```{r}
ggplot(gapminder,aes(x=year,y=lifeExp))+
  geom_point(color='red',alpha=0.5)+
  geom_smooth(se=FALSE)+
  facet_wrap(~continent,nrow=2)+
  scale_x_continuous(breaks = seq(1950, 2010, 10),
                     labels = as.character(seq(1950, 2010, 10)),
                     limits = c(1950, 2010))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size=14,hjust=0.5))+
  labs(x="year",
       y="life expectancy",
       title="life expectancy of all continents over time")
```

## Report the absolute abundance of countries with low life expectancy over time by continent

First we need to compute worldwide low life expectancy for each year. low life expectancy = average life expectancy - standard deviation of life expectancy. 

```{r}
group_by(gapminder,year)%>%
  summarize(low_life_expectancy=mean(lifeExp)-sd(lifeExp))%>%kable()
```

Next I will compute how many countries on each continent have a life expectancy less than low life expectancy. 

```{r}
f<-group_by(gapminder,year)%>%
  mutate(lowLifeExp=mean(lifeExp)-sd(lifeExp)) #determine low life expectancy for each year
g<-group_by(f,continent,year)%>%
  summarize(numBelowLowLifeExp=sum(lifeExp<lowLifeExp))
filter(g,continent == "Africa")%>%kable() #this produces a long table, I will only output data for Africa as an example
```

Finally I will plot this using scatter plot and line plot.

```{r}
ggplot(g,aes(x=year,y=numBelowLowLifeExp))+
  geom_point(color='red',alpha=0.5)+
  geom_line()+
  facet_wrap(~continent)+
  scale_x_continuous(breaks = seq(1950, 2010, 10),
                     labels = as.character(seq(1950, 2010, 10)),
                     limits = c(1950, 2010))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size=14,hjust=0.5))+
  labs(x="year",
       y="number of countries below low life expectancy",
       title="Absolute abundance of countries with low life expectancy over time")
```

We can find that Africa has the most countries with low life expectancy, and it even has an increasing tendency over time.
  
  

  
